--- 
title: "Cyclistic Case Study"
author: "Dwayne Marson"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
setwd("~/DataAnalyticsCapstone/")
```


### **Cyclistic Bikeshare Analysis**

#### **An optional project done as part of the Google Dta Analytics Certificate course.**

#### **Preparing the Environment**

```{r}
# Loading the tools needed for the analysis

library(tidyverse) # For data manipulation and plotting
library(readr) # For reading dataset into the environment
library(dplyr)
library(lubridate) # For handling dates and time
library(janitor) # For cleaning column names
library(ggplot2)
library(scales)
library(knitr)
```

### **2. Preparing The Data for Processing**

Imported all datasets and renamed them for better workflow.

```{r}

# Import all Datasets

df_202404_tripdata <- read_csv("data/202404-divvy-tripdata.csv")
df_202405_tripdata <- read_csv("data/202405-divvy-tripdata.csv")
df_202406_tripdata <- read_csv("data/202406-divvy-tripdata.csv")
df_202407_tripdata <- read_csv("data/202407-divvy-tripdata.csv")
df_202408_tripdata <- read_csv("data/202408-divvy-tripdata.csv")
df_202409_tripdata <- read_csv("data/202409-divvy-tripdata.csv")
df_202410_tripdata <- read_csv("data/202410-divvy-tripdata.csv")
df_202411_tripdata <- read_csv("data/202411-divvy-tripdata.csv")
df_202412_tripdata <- read_csv("data/202412-divvy-tripdata.csv")
df_202501_tripdata <- read_csv("data/202501-divvy-tripdata.csv")
df_202502_tripdata <- read_csv("data/202502-divvy-tripdata.csv")
df_202503_tripdata <- read_csv("data/202503-divvy-tripdata.csv")


```

#### **Inspecting table structures**

```{r}
# Comparing the columns in each of the files

colnames(df_202404_tripdata)
colnames(df_202405_tripdata)
colnames(df_202406_tripdata)
colnames(df_202407_tripdata)
colnames(df_202408_tripdata)
colnames(df_202409_tripdata)
colnames(df_202410_tripdata)
colnames(df_202411_tripdata)
colnames(df_202412_tripdata)
colnames(df_202501_tripdata)
colnames(df_202502_tripdata)
colnames(df_202503_tripdata)

# Take a look at the structure of each table, taking note of the data type in each column

str(df_202404_tripdata)
str(df_202405_tripdata)
str(df_202406_tripdata)
str(df_202407_tripdata)
str(df_202408_tripdata)
str(df_202409_tripdata)
str(df_202410_tripdata)
str(df_202411_tripdata)
str(df_202412_tripdata)
str(df_202501_tripdata)
str(df_202502_tripdata)
str(df_202503_tripdata)
```

### **3. Processing the Data**

```{r}
# The 12 datasets are then merged into a single data table.

Divvy_12_Months_data <- bind_rows(df_202404_tripdata, df_202405_tripdata, df_202406_tripdata,
                                  df_202407_tripdata, df_202408_tripdata, df_202409_tripdata, 
                                  df_202410_tripdata, df_202411_tripdata, df_202412_tripdata,
                                  df_202501_tripdata, df_202502_tripdata, df_202503_tripdata)
```

#### **Renamed a column to ensure easy reading.**

```{r}
# Rename the column member_casual to rider_type as this makes it more readable

Divvy_12_Months_data <- Divvy_12_Months_data %>% 
  rename(user_type = member_casual,
         trip_id = ride_id,
         bike_type = rideable_type,
         start_time = started_at,
         end_time = ended_at,
         )
```

#### **Created 4 new columns to allow for easier access to information needed for the analysis.(ride_length, day_of_the_week, month, date)**

```{r}

  
# 1. ride length" in minutes to determine how long riders use the bikes for and a column for day of the week

Divvy_12_Months_data <- Divvy_12_Months_data %>% 
  mutate(
    ride_length = as.numeric(difftime(end_time, start_time, units = "mins")),
    day_of_week = wday(start_time,label = TRUE, abbr = FALSE) # this create the full name of the day of the week
  )

# 2. creating a month and date column to help with seasonal trends

Divvy_12_Months_data <- Divvy_12_Months_data %>% 
  mutate(month = month(start_time, label = TRUE, abbr = FALSE))
Divvy_12_Months_data <- Divvy_12_Months_data %>%  
  mutate(date = as.Date(start_time))
```

#### **Cleaning data**

some fields were filtered out. These include :

```{r}
#Rides with negative ride length and rides over 24 hours
Divvy_12_Months_data <- Divvy_12_Months_data %>% 
  filter(ride_length > 0, ride_length < 1440) %>% 
  drop_na() 

# this keeps the rides between 1 minute and 24 hrs


# Removing duplicate rows
Divvy_12_Months_data <- distinct(Divvy_12_Months_data)

# removing rows will null value
Divvy_12_Months_data <- na.omit(Divvy_12_Months_data)
```

### **4. Data Analysis**

The analysis focuses on identifying key differences in ride behavior
between **casual riders** and **annual members** across four main
dimensions:\
- Ride length\
- Day-of-week usage\
- Ride duration by day\
- Monthly trends

#### **Ride Length Summary**

```{r}
# Summary statistics for ride length (average, median, max and min ride length)

Divvy_12_Months_data %>% 
  group_by(user_type) %>% 
  summarize(
    `Mean Ride Length` = round(mean(ride_length), 2),
    `Median Ride Length` = round(median(ride_length), 2),
    `Max Ride Length` = round(max(ride_length), 2),
    `Min Ride Length` = round(min(ride_length), 2),
    .groups = "drop"
  ) %>%
  rename(`User Type` = user_type) %>%
  kable(
    caption = "Summary Statistics for Ride Length by Rider Type",
    align = "lcccc"
  )
```

**Summary**

Casual riders have significantly **longer average and median ride
durations** than members.

Both groups include rides up to 24 hours, likely due to **return errors
or outliers**, but casual riders consistently average longer durations.

This suggests **casual riders are more leisure-oriented**, while
**members use the service more efficiently**, likely for commuting or
daily errands.

#### **Ride Frequency by Day of Week**

```{r}
# How ride behaviours change during the week
Divvy_12_Months_data %>% 
  group_by(user_type, day_of_week) %>% 
  summarize(number_of_rides = n()) %>% 
 # arrange(rider_type, day_of_week) %>% 
  pivot_wider(
    names_from = user_type,
    values_from = number_of_rides
  ) %>%
  rename(
    Day = day_of_week,
    `Casual Rider` = casual,
    `Member Rider` = member
  ) %>%
  kable(
    caption = "Ride Frequency By Day of the Week",
    col.names = c("Day", "Casual Riders", "Member Riders"),
    align = "lcc"
  )
  
```

-   **Members ride more frequently on weekdays**, peaking midweek
    (Wednesday).\
-   **Casual riders are most active on weekends**, especially Saturdays.

This supports the hypothesis that **members ride for commuting**, while
**casual riders use bikes more for recreation or tourism**.

------------------------------------------------------------------------

#### **Average Ride Duration by Day of Week**

```{r}
# Average ride length per day

# Generate table
Divvy_12_Months_data %>%
  group_by(user_type, day_of_week) %>%
  summarize(average_ride_length = mean(ride_length), .groups = "drop") %>%
  mutate(average_ride_length = round(average_ride_length, 1)) %>%
  pivot_wider(
    names_from = user_type,
    values_from = average_ride_length
  ) %>%
  rename(
    Day = day_of_week,
    `Casual Rider` = casual,
    `Member Rider` = member
  ) %>%
  kable(
    caption = "Average Ride Length by Day (in minutes)",
    col.names = c("Day", "Casual Riders (Avg min)", "Member Riders (Avg Min)"),
    align = "lcc"
  )
```

-   Casual riders consistently spend nearly **twice as long per ride**
    on all days.\
-   **Weekend rides are longest for both**, especially for casuals (27+
    minutes).\
-   Members maintain **short, stable ride durations** — reinforcing
    commuter use.

------------------------------------------------------------------------

#### **Monthly Ride Patterns**

```{r}
# Seasonal trends

Divvy_12_Months_data %>% 
  group_by(user_type, month) %>% 
  summarize(number_of_rides = n(), .groups = "drop") %>% 
  pivot_wider(
    names_from = user_type,
    values_from = number_of_rides
  ) %>%
  rename(
    Month = month,
    `Casual Rider` = casual,
    `Member Rider` = member
  ) %>%
  kable(
    caption = "Monthly Ride Counts by Rider Type",
    col.names = c("Month", "Casual Rider", "Member Rider"),
    align = "lcc"
  )
```

#### **Casual Riders**

-   Peaks in **July (231,963)** and **August (228,502)**\
-   Lowest in **January (17,090)** and **February (19,631)**

#### **Members**

-   Peaks in **September (320,853)** and **August (312,803)**\
-   Still strong in **January (84,121)** and **February (89,945)**

**Insights**: - Casual usage is **highly seasonal**, with steep summer
peaks.\
- Member usage is **more consistent year-round**, supporting regular,
utility-driven behavior.\
- This reinforces the idea that **membership is better suited to
frequent, all-weather users**.

------------------------------------------------------------------------

#### **Key Behavioral Differences**

| Category            | Casual Riders            | Members                 |
|---------------------|--------------------------|-------------------------|
| Purpose             | Leisure, tourism         | Commuting, errands      |
| Ride duration       | Longer (avg. 23.8 min)   | Shorter (avg. 12.4 min) |
| Ride frequency      | High on weekends         | High on weekdays        |
| Seasonality         | Summer peak, winter drop | Year-round, steady      |
| Day with most rides | Saturday (316,461)       | Wednesday (424,492)     |

------------------------------------------------------------------------

### **5. Sharing the Data through visualization**

#### **Plotting the charts**

```{r}

#order the days of the week ensuring the days are listed in order from Sunday to Saturday

day_order <- c("Sunday", "Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday")

# Number of rides by day of the week
####
ride_by_weekday <- Divvy_12_Months_data %>% 
  mutate(day_of_week = factor(day_of_week, levels = day_order)) %>% 
  group_by(user_type, day_of_week) %>% 
  summarise(number_of_rides = n(), .groups = "drop") %>% 
  ggplot(aes(x = day_of_week, y = number_of_rides, fill = user_type)) +
  geom_col(position = position_dodge(width = 0.9)) +
  geom_text(
    aes(label = comma(number_of_rides)),
    position = position_dodge(width = 0.9),
    vjust = -0.5,
    size = 3
  ) +
  scale_y_continuous(labels = comma) +
  labs(
    title = "Number of Rides by Weekday",
    x = "Day of Week",
    y = "Number of Rides",
    fill = "Member Type"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5),
    plot.caption = element_text(hjust = 0.5),
    legend.position = "top",
    legend.justification = "center",
    axis.text.x = element_text(angle = 0, hjust = 1))
# Display the plot
ride_by_weekday
```

#### **Average ride length by day of the week**

```{r}
#average ride length by day of the week
#####
Avg_ride_length_by_weekday <- Divvy_12_Months_data %>% 
  mutate(day_of_week = factor(day_of_week, levels = day_order)) %>%
  group_by(user_type, day_of_week) %>% 
  summarise(average_ride_length = mean(ride_length), .groups = "drop") %>% 
  ggplot(aes(x = day_of_week, y = average_ride_length, fill = user_type)) +
  geom_col(position = position_dodge(width = 0.9)) +
  geom_text(
    aes(label = round(average_ride_length, 1)),  # or use comma(average_ride_length)
    position = position_dodge(width = 0.9),
    vjust = -0.5,
    size = 3
  ) +
  scale_y_continuous(labels = comma) +
  labs(
    title = "Average Ride Length by Weekday",
    x = "Day of the Week",
    y = "Average Ride Length (min)",
    fill = "Member Type"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5),
    plot.caption = element_text(hjust = 0.5),
    legend.position = "top",
    legend.justification = "center",
    axis.text.x = element_text(angle = 0, hjust = 1))
# Display the plot
Avg_ride_length_by_weekday
```

#### **Number of rides by types of bike**

```{r}
#Number of rides by types of bike
####
rides_by_type_of_bike <- Divvy_12_Months_data %>% 
  group_by(user_type, bike_type) %>% 
  summarise(number_of_rides = n(), .groups = "drop") %>% 
  ggplot(aes(x = bike_type, y = number_of_rides, fill = user_type)) +
  geom_col(position = position_dodge(width = 0.9)) +
  geom_text(
    aes(label = comma(number_of_rides)),
    position = position_dodge(width = 0.9),
    vjust = -0.5,
    size = 3
  ) +
  scale_y_continuous(labels = comma) +
  labs(
    title = "Bike Type Usage by Rider Type",
    x = "Bike Type",
    y = "Number of Rides",
    fill = "User Type"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5),
    plot.caption = element_text(hjust = 0.5),
    legend.position = "top",
    legend.justification = "center",
    axis.text.x = element_text(angle = 0, hjust = 1))
# Display the plot
rides_by_type_of_bike
```

#### **Rides per month by rider type**

```{r}
# defining the month order

month_order <- c("January", "February", "March", "April", "May", "June", "July", "August", "September",
"October", "November", "December")

#Rides per month by rider type
###
rides_per_month_by_ridertype <- Divvy_12_Months_data %>%
  mutate(month = factor(month, levels = month_order)) %>% 
  group_by(user_type, month) %>% 
  summarise(number_of_rides = n(), .groups = "drop") %>% 
  ggplot(aes(x = month, y = number_of_rides, fill = user_type)) +
  geom_col(position = position_dodge(width = 0.9)) +
  geom_text(
    aes(label = comma(number_of_rides)),
    position = position_dodge(width = 0.9),
    vjust = -0.5,
    size = 3
  ) +
  scale_y_continuous(labels = comma) +
  labs(
    title = "Number of Rides per Month By Ridertype",
    x = "Month",
    y = "Number of Rides",
    fill = "User Type"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5),
    plot.caption = element_text(hjust = 0.5),
    legend.position = "top",
    legend.justification = "center",
    axis.text.x = element_text(angle = 45, hjust = 1))
# Display the plot
rides_per_month_by_ridertype
```

#### **Top 10 Peak Usage Days for Casual Riders**

```{r}
# Step 1: Get daily ride counts for casual riders
casual_daily_counts <- Divvy_12_Months_data %>%
  filter(user_type == "casual") %>%
  group_by(date) %>%
  summarise(rides = n(), .groups = "drop")

# Step 2: Get top 10 highest-usage days
top_10_dates <- casual_daily_counts %>%
  arrange(desc(rides)) %>%
  slice_head(n = 10) %>%
  mutate(date_label = format(date, "%B %d, %Y"))  # for cleaner x-axis labels

# Step 3: Bar chart with labels and rotated x-axis
ggplot(top_10_dates, aes(x = reorder(date_label, -rides), y = rides)) +
  geom_col(fill = "steelblue") +
  geom_text(
    aes(label = scales::comma(rides)),
    vjust = -0.5,
    size = 3
  ) +
  labs(
    title = "Top 10 Peak Usage Days for Casual Riders",
    x = "Date of Rides",
    y = "Number of Rides"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(hjust = 0.5, size = 14, face = "bold")
  )
```

### **6. Acting on the findings of the analysis**

#### **Recommendations**

Based on the analysis, I am recommending three actionable strategies

**1. Target casual riders with weekday commuter incentives** Insight:
Casual riders take longer rides mostly on weekends, while members ride
shorter distances but more consistently on weekdays.

-   Promote the time and cost savings of membership for daily use.
-   Launch campaigns encouraging casual riders to try weekday commuting

**2. Offer seasonal promotions or loyalty discounts to casual riders**
Insight: Casual ridership spikes in summer months between June and
September, while members ride more evenly year round. - Especially
during the months when casual ridership is highest. - Tailor messages
around recreational convenience plus long-term savings.

**3. Highlight the value of membership through rideable type Access and
cost** Insight: Casual riders use electric and docked bikes more often,
which incur higher per ride cost - Promote the financial benefits of
becoming a member

#### **Conclusion**:

There are clear behavioral distinctions between casual and member
riders. Members show consistent, short-duration, weekday usage — typical
of commuter habits. Casual riders exhibit longer, leisure-style rides
that spike on weekends and in summer. These insights provide a strong
foundation for marketing strategies targeting casual-to-member
conversion.
